Principal Component Analysis (PCA)
====================
We will use the `USArrests` data (which is in R)

`USArrests` data is a data frame with 50 observations on 4 variables.

[,1]	Murder (Numeric):	Murder arrests (per 100,000)

[,2]	Assault (Numeric):Assault arrests (per 100,000)

[,3]	UrbanPop (Numeric):	Percent urban population

[,4]	Rape (Numeric): Rape arrests (per 100,000)


```{r}
head(USArrests)
dimnames(USArrests)
```

The *apply()* function allows us to apply a formula. Let's use this to calculate the mean and variance.

Note: The second input here denotes whether we wish to compute the mean of the rows (1) or the columns (2).


```{r}
apply(USArrests,2,mean)
apply(USArrests,2, var)
```

We see that `Assault` has a much larger variance than the other variables. It would dominate the principal components, so we choose to **standardize** the variables to have mean 0 and SD 1 when we perform PCA.


We now perform Principal Components Analysis using the **prcomp()** function

```{r}
pca.out=prcomp(USArrests, scale=TRUE)
```

Let's examine the **pca.out** object. Some observations:

* Standard Deviation is the SD of the Principal Components. NOTE how they are all decreasing.

* Rotation refers to the loadings. e.g. First principal component is loaded almost equally across all compnents (like an average across all 3).

* PCA doesn't care about the signs - all about the variance

* First Principal component focuses on crime

* Second principal component focuses on urban pop
```{r}
pca.out
```

Let's get a summary of the object

```{r}
summary(pca.out)
```


Now, let's look at a scree plot. This plots the variance explained by the
principal components.

```{r}
layout(matrix(1:2, ncol=2))
screeplot(pca.out)
screeplot(pca.out, type="lines")
```

Plot the "Percentage of variance Explained" (PVE) of each principal component (i.e. a scree plot) and the cumulative PVE of each principal component. This can be done with just a
little work.

```{r}
pve=100*pca.out$sdev^2/sum(pca.out$sdev^2)
par(mfrow=c(1,2))
plot(pve, type="o", ylab="PVE", xlab="Principal Component",col =" blue")
plot(cumsum(pve), type="o", ylab="Cumulative PVE", xlab="Principal Component ", col =" brown3 ")
```


Finally, let's plot the first two principal components and analyse the output

* Biplot: Both the loadings/ principal component scores and the observations (i.e. states) are in the one plot
* Observations are the states
* Red shows the directions of the loadings for the principal components. 
* First Axis - largely related to the amount of crime in the state
* Second Axis - loading is mainly urban population, so this axis is whether state has large urban population or not.


```{r}
biplot(pca.out, scale=0, cex = 0.6)
```

